Think deep for reef resilience

### 2025-08-16 paper/paper_draft.qmd
Standardizing text
low medium high -> Low Medium High
management area -> Management area

clusters in "low clusters" -> categories
The above is to avoid confusion:
- Reef clusters, time series clusters

features -> reef characteristics
This is to avoid confusion between reef features (their spatial definitions) and features in the sense of inputs/factors of an ML model

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority (GBRMPA)
[@GBRMPA2022]. The reef definitions define 3806 reefs across the GBR. An indicative map of
the GBR marine park including the Marine Park management areas [@GBRMPA2007] and the
corresponding bioregion for each reef is shown in (@fig-region-map; [@GBRMPA2024]).
Estimates of coral carrying capacity for each reef were taken from prior analyses undertaken
in [@bozec2022].

@@

While the bathymetry data adopted for the given assessment are said to be accurate up to 15m
from Mean Sea Level [@GBRMPA2020], there are many sources of uncertainty in the depth
estimates used for this study. These are outlined further in the Discussion. Notably, the
adopted depth estimates were derived from bathymetries based on satellite imagery. The
bathymetry data can be limited in their alignment with the established GBRMPA reef outlines
used to characterise reef locations and their areal extent. A Quality Assessment and Control
process identified reefs with potential data limitations: reefs with no overlapping
bathymetry data, reefs with minimum depth values above sea level (indicating potential data
artifacts), and reefs where bathymetry pixels covered less than 5% of the reef polygon area.
This process removed 842 reefs, or 22%, from the initial set of 3806 reefs. Furthermore,
reefs that are not associated with GBRMPA management areas [@GBRMPA2007] or reefal
bioregions [@GBRMPA2024] were also removed from the investigation. Bioregions which
contained fewer than three reefs were also excluded. As a result, 1377 reefs were excluded
(36%), leaving a total of 2429 remaining for the purpose of this investigation.

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points). Colour-coded areas outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2007]. Reef centroids are marked individually for visual clarity, and are coloured
according to its bioregion [@GBRMPA2024]. All 2429 reefs included in this investigation are
represented in this figure. Annotations represent major coastal settlements along the
GBR.](../figs/region_map.png){#fig-region-map fig-align="center" width="14.8cm"}

@@

![Degree Heating Week time series (in °C - Weeks) for the GCMs selected. Colours represent
the different management areas across the Great Barrier Reef. Green represents
Cairns-Cooktown, orange represents Far-Northern, blue represents Mackay-Capricorn and red
represents the Townsville-Whitsunday management area. Solid lines represent the
management-area median time series and bands represent the 95th percentile confidence
intervals.](../figs/methods_dhw_timeseries.png){#fig-methods-dhws fig-align="center"
width="14.82cm"}

@@

The time series clustering was applied over the time frame between 2030 and 2060. The
selected time frame prevents the classification from being overly influenced by any steep
declines in the initial years of simulations, and avoids the post-2060 time period where
reefs are projected to be under severe stress and in decline [@bozec2025]. Spatially, the
analyses were conducted for each GBR management area and at the scale of each reefal
bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical results **\[TI: Ben, compared to
what, the bioregion or the management area results?\]** and so are not shown here. To
compare time series clustering results across GCMs, management area scale clustering results
are used to preserve coarse scale latitudinal differences between regions, while maintaining
ease of comparison and visualisation. These comparisons include quantification of
differences between low and high cover clusters across management regions under each GCM
using the CID between cluster median time series. To assess how different reef
characteristics influence time series clustering at a finer spatial scale, bioregion results
are used to capture patterns occurring within individual clusters of reefs. Results of all
analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for those
interested.

@@

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median timeseries in each management area under each GCM. CID values are calculated for percentage cover timeseries (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) timeseries (@sfig-gcm-dhw-trajectories). Complexity invariant distance represents the euclidean distance between timeseries, corrected by the complexity of the two timeseries. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW timeseries are not directly comparable as the timeseries use different units.

@@

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to
2075. The assessed carbonate budget thresholds range from 10 to 20% of total reef area.
Bracketed x-axis values represent the Spearman's correlation between reef depth and the
number of years cover exceeds the carbonate budget threshold. Point colour represents the
median reef depth level in metres. Displays results from median timeseries using
ACCESS-ESM1-5.](../figs/ACCESS-ESM1-5/depth_carbonate_scatter.png){#fig-access-esm1-5-depth-carbonate
fig-align="center" width="14.82cm"}

@@

Under all GCMs many reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels, with an average of 38% of reefs (924 reefs) clustered into the
lowest cover level cluster at the management area scale across GCMs. NorESM2-MM and
ACCESS-ESM1-5 GCMs convey the most coral cover maintenance into the future. Differences
between low, medium and high reef clusters in cover timeseries are most notable under these
two GCMs [@tbl-cluster-cid; @fig-gcm-cover-trajectories]. Maintenance of coral cover into
the future is linked to the Equilibrium Climate Sensitivity of the GCM used as levels of
future recovery and timespan of cover maintenance increase as GCM ECS increases
[@fig-gcm-cover-trajectories]. Under high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2)
large declines and lower rates of recovery in high cover clusters are common compared to
lower-ECS GCMs.

The large differences in cover timeseries seen across the GCMs suggest that it is important
to consider uncertainty in future conditions when assessing possible biological responses to
change. This is supported by recent literature suggesting an effective approach to capturing
climate change uncertainty is to model responses using GCMs separately [@shepherd2018].
Despite large differences in coral cover across GCMs, the patterns identified between low,
medium and high clusters of reefs suggest that the relative importance of environmental and
biological factors such as depth refugia, connectivity and surface DHW levels is consistent
across scenarios.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Our results suggest that within shallow-water reef ecosystems
(non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability
of the reef to maintain positive carbonate budgets under future conditions. While similar
findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses
have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific. Reefs at shallow depths experience no, or few, years with
coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative carbonate budgets in the future may contribute to
phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting
algae driven production, or to bioerosion dominated systems with rapid erosion of physical
reef frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer.

>>>>

The spatial extent of the GBR reefs represented in this investigation follows the reef
definitions available via the Great Barrier Reef Marine Park Authority (GBRMPA)
[@GBRMPA2022]. The reef definitions define 3806 reefs across the GBR. An indicative map of
the GBR marine park including the Marine Park Management areas [@GBRMPA2007] and the
corresponding bioregion for each reef is shown in (@fig-region-map; [@GBRMPA2024]).
Estimates of coral carrying capacity for each reef were taken from prior analyses undertaken
in [@bozec2022].

@@

While the bathymetry data adopted for the given assessment are said to be accurate up to 15m
from Mean Sea Level [@GBRMPA2020], there are many sources of uncertainty in the depth
estimates used for this study. These are outlined further in the Discussion. Notably, the
adopted depth estimates were derived from bathymetries based on satellite imagery. The
bathymetry data can be limited in their alignment with the established GBRMPA reef outlines
used to characterise reef locations and their areal extent. A Quality Assessment and Control
process identified reefs with potential data limitations: reefs with no overlapping
bathymetry data, reefs with minimum depth values above sea level (indicating potential data
artifacts), and reefs where bathymetry pixels covered less than 5% of the reef polygon area.
This process removed 842 reefs, or 22%, from the initial set of 3806 reefs. Furthermore,
reefs that are not associated with GBRMPA Management areas [@GBRMPA2007] or reefal
bioregions [@GBRMPA2024] were also removed from the investigation. Bioregions which
contained fewer than three reefs were also excluded. As a result, 1377 reefs were excluded
(36%), leaving a total of 2429 remaining for the purpose of this investigation.

![Map of the Great Barrier Reef region, including GBRMPA Management areas (polygons) and all
considered reefs (points). Colour-coded areas outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2007]. Reef centroids are marked individually for visual clarity, and are coloured
according to its bioregion [@GBRMPA2024]. All 2429 reefs included in this investigation are
represented in this figure. Annotations represent major coastal settlements along the
GBR.](../figs/region_map.png){#fig-region-map fig-align="center" width="14.8cm"}

@@

![Degree Heating Week time series (in °C - Weeks) for the GCMs selected. Colours represent
the different Management areas across the Great Barrier Reef. Green represents
Cairns-Cooktown, orange represents Far-Northern, blue represents Mackay-Capricorn and red
represents the Townsville-Whitsunday Management area. Solid lines represent the
management-area median time series and bands represent the 95th percentile confidence
intervals.](../figs/methods_dhw_timeseries.png){#fig-methods-dhws fig-align="center"
width="14.82cm"}

@@

The time series clustering was applied over the time frame between 2030 and 2060. The
selected time frame prevents the classification from being overly influenced by any steep
declines in the initial years of simulations, and avoids the post-2060 time period where
reefs are projected to be under severe stress and in decline [@bozec2025]. Spatially, the
analyses were conducted for each GBR Management area and at the scale of each reefal
bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical results **\[TI: Ben, compared to
what, the bioregion or the Management area results?\]** and so are not shown here. To
compare time series clustering results across GCMs, Management area scale clustering results
are used to preserve coarse scale latitudinal differences between regions, while maintaining
ease of comparison and visualisation. These comparisons include quantification of
differences between low and high cover clusters across management regions under each GCM
using the CID between cluster median time series. To assess how different reef
characteristics influence time series clustering at a finer spatial scale, bioregion results
are used to capture patterns occurring within individual clusters of reefs. Results of all
analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for those
interested.

@@

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median time series in each management area under each GCM. CID values are calculated for percentage cover time series (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) time series (@sfig-gcm-dhw-trajectories). Complexity invariant distance represents the euclidean distance between time series, corrected by the complexity of the two time series. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW time series are not directly comparable as the time series use different units.

@@

![Number of years reef coral cover exceeds the carbonate budget threshold level from 2025 to
2075. The assessed carbonate budget thresholds range from 10 to 20% of total reef area.
Bracketed x-axis values represent the Spearman's correlation between reef depth and the
number of years cover exceeds the carbonate budget threshold. Point colour represents the
median reef depth level in metres. Displays results from median time series using
ACCESS-ESM1-5.](../figs/ACCESS-ESM1-5/depth_carbonate_scatter.png){#fig-access-esm1-5-depth-carbonate
fig-align="center" width="14.82cm"}

@@

Under all GCMs many reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels, with an average of 38% of reefs (924 reefs) clustered into the
lowest cover level cluster at the management area scale across GCMs. NorESM2-MM and
ACCESS-ESM1-5 GCMs convey the most coral cover maintenance into the future. Differences
between Low, Medium and High reef clusters in cover time series are most notable under these
two GCMs [@tbl-cluster-cid; @fig-gcm-cover-trajectories]. Maintenance of coral cover into
the future is linked to the Equilibrium Climate Sensitivity of the GCM used as levels of
future recovery and timespan of cover maintenance increase as GCM ECS increases
[@fig-gcm-cover-trajectories]. Under high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2)
large declines and lower rates of recovery in high cover clusters are common compared to
lower-ECS GCMs.

The large differences in cover time series seen across the GCMs suggest that it is important
to consider uncertainty in future conditions when assessing possible biological responses to
change. This is supported by recent literature suggesting an effective approach to capturing
climate change uncertainty is to model responses using GCMs separately [@shepherd2018].
Despite large differences in coral cover across GCMs, the patterns identified between low,
medium and high clusters of reefs suggest that the relative importance of environmental and
biological factors such as depth refugia, connectivity and surface DHW levels is consistent
across scenarios.

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced,
and exhibit overall similar DHW time series. While these reefs experience similar magnitude
DHWs based on SST, the proportion of the reefs' coral population impacted by heat stress
events depends on reef depth. Our results suggest that within shallow-water reef ecosystems
(non-mesophotic, depth \< 30m), depth contributes greatly to reef resilience and the ability
of the reef to maintain positive carbonate budgets under future conditions. While similar
findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses
have often focussed on the potential of mesophotic reefs (\> 30m deep) as thermal refugia
[@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019]
found no association between depth and the alleviation of heat stress across a range of
shallow reefs in the Pacific. Reefs at shallow depths experience no, or few, years with
coral cover exceeding levels likely required to maintain positive carbonate budgets.
Movement of shallow reefs towards negative carbonate budgets in the future may contribute to
phase shifts in reef dominant CaCO~3~ production. Shifting from coral driven to encrusting
algae driven production, or to bioerosion dominated systems with rapid erosion of physical
reef frameworks [@dudgeon2010; @perry2008]. While reefs at increased depths may be able to
maintain coral driven CaCO~3~ production and accumulation into the future for longer.
